---
title: "Dataset download for Cherry Trees Prediction "
author: "Ha Dao"
date: "01/01/2023"
output:
  html_document:
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, error = FALSE, 
                      message = FALSE,
                      fig.align = 'center',
                      out.width = '80%')
```

- Data Exploration and Visualizing (intial)
```{r}
library(tidyverse)
library(rnoaa)     # Weather data
library(imputeTS)  # Imputation on the temperature
library(lubridate)
library(randomForest) 
library(e1071)     # SVM
library(ggsci)     # Color
library(corrplot)  # Correlation plot
library(hash)
library(grid)
library(ggnewscale)
library(ggtext)
library(tidyverse)

```

## Bloom data from 3 main locations: 
```{r}
# Blossom data from 3 main locations: 
cherry <- read.csv("peak-bloom-prediction-main-2/data/washingtondc.csv") %>% 
  bind_rows(read.csv("peak-bloom-prediction-main-2/data/liestal.csv")) %>% 
  bind_rows(read.csv("peak-bloom-prediction-main-2/data/kyoto.csv")) %>%
  bind_rows(read.csv("peak-bloom-prediction-main-2/data/vancouver.csv")) 
```

Bloom date data from Liestal (since we can't find data for Vancouver): 
```{r}
# We will try to use K-means to see wheter weather from Vancouver is most similar to which country  

# https://highparknaturecentre.com/index.php/cherry-blossom-bloom-tracking
# Peak Bloom date from Liestal ( might be better if I can find data for Cali bloomdate)
cherry_vc <- cherry %>%
  filter(location %in% c("liestal")) %>%
  select(year, bloom_date, bloom_doy)
         
# cherry_vc <- cherry_vc %>%
#   #mutate(bloom_date = as.POSIXct(bloom_date, format = '%m/%d/%Y')) %>%
#   #mutate(year = as.integer(format(bloom_date, "%Y"))) %>%
#   #mutate(bloom_doy = yday(bloom_date)) %>%
#   mutate(location = "vancouver")
#   #mutate(bloom_date = as.character(bloom_date))
#   #select(location, year, bloom_date, bloom_doy)
# 
# cherry_vc <- cherry_vc %>%
#   filter(year %in% c(2006:2023))
# 
# 
# cherry <- cherry %>% select(location, year, bloom_date, bloom_doy) %>%
#   bind_rows(cherry_vc)


```


```{r}
cherry %>% 
  group_by(location) %>% 
  slice_head(n = 4) 
```
## Visualizing the time series
```{r, fig.width=8, fig.height=3, out.width='100%', fig.cap="Time series of peak bloom of cherry trees since 1880 at three different sites."}
cherry %>% 
  filter(year >= 1960) %>%
  ggplot(aes(x = year, y = bloom_doy)) +
  geom_point() +
  geom_step(linetype = 'dotted', color = 'gray50') +
  scale_x_continuous(breaks = seq(1960, 2023, by = 20)) +
  scale_y_continuous(breaks = seq(60, 150, by = 5)) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(x = "Year", y = "Peak bloom (days since Jan 1st)")
```


```{r, fig.cap="Distribution of peak bloom days across the four sites."}
cherry %>%
  ggplot() +
  geom_density(aes(x = bloom_doy, fill = location), alpha = 0.7) +
  scale_fill_npg() + theme_light()+
  labs(x = "Peak bloom (days since Jan 1st)", y = "Density")
```



1. Get the weather data from RNOAA

The list of stations can be retrieved using the `ghcnd_stations()` function. Note that the closest weather station to each city with continuously collected maximum temperatures are USC00186350 (Washington D.C.), GME00127786 (Liestal), JA000047759 (Kyoto), and CA001108395 (Vancouver).

```{r, eval = FALSE}
stations <- ghcnd_stations()
```

As a simple demonstration, we retrieve the average seasonal maximum daily temperature (in 1/10 °C) from these stations using our own `get_temperature()` function, which wraps the `ghcnd_search()` function in the `rnoaa` package. (N.b. `ghcnd_search()` returns a list. Each element of the list corresponds to an element of the `var` argument.)

```{r}
#' Get the annual average maximum temperature at the given station,
#' separated into the 4 meteorological seasons (Winter, Spring, Summer, Fall).

#' The seasons are span 3 months each.
#' Winter is from December to February, Spring from March to May,
#' Summer from June to August, and Fall from September to November.
#' Note that December is counted towards the Winter of the next year, i.e.,
#' temperatures in December 2020 are accounted for in Winter 2021.

#' @param stationid the `rnoaa` station id (see [ghcnd_stations()])
#' @return a data frame with columns
#'   - `year` ... the year of the observations
#'   - `season` ... the season (Winter, Spring, Summer, Fall)
#'   - `tmax_avg` ... average maximum temperate in tenth degree Celsius
# Define function to get weather data for a specific station and date range
get_temperature <- function(station_id = station_id, date_range = c("1930-01-01", "2024-03-30")) {
  
  # Get daily maximum temperature data, minimum temperature and precipitation data
  tmax <- ghcnd_search(stationid = station_id, var = c("tmax"), 
                        date_min = date_range[1], date_max = date_range[2])[[1]]
  tmin <- ghcnd_search(stationid = station_id, var = c("tmin"), 
                        date_min = date_range[1], date_max = date_range[2])[[1]]
  prcp <- ghcnd_search(stationid = station_id, var = c("prcp"), 
                        date_min = date_range[1], date_max = date_range[2])[[1]]
  # Merge with tmax
  weather <- merge(tmax, tmin, by = "date") %>% merge(prcp, by = "date")
  
  # Select and transform relevant columns
  weather <- weather %>% 
    select(date, tmax, tmin, prcp) %>%
    mutate(tmax = tmax/10, tmin = tmin/10, # Convert temperature to tenth of the C
           year = as.integer(format(date, "%Y")),
           month = as.integer(strftime(date, '%m')) %% 12, # Make December "0"
           season = cut(month, breaks = c(0, 2, 5, 8, 11),
                        include.lowest = TRUE,
                        labels = c("Winter", "Spring", "Summer", "Fall")),
           year = if_else(month == 0, year + 1L, year)) %>%
    mutate(month_name = if_else(month == 0, 12, month)) %>%
    mutate(month_name = factor(month.name[as.integer(month_name)], levels = month.name)) %>%
    group_by(year, month, season) %>%
    summarize(tmax_avg = mean(tmax, na.rm = TRUE),
              tmax_max = max(tmax, na.rm = TRUE),
              tmin_avg = mean(tmin, na.rm = TRUE),
              tmin_min = min(tmin, na.rm = TRUE),
              prcp_avg = mean(prcp, na.rm = TRUE),
              avg_temp = (tmax_avg + tmin_avg)/2)

  # Return transformed data
  return(weather)
}
```

```{r}
#Get data
historic_temperatures <-
  tibble(location = "washingtondc", get_temperature("USC00186350")) %>%
  bind_rows(tibble(location = "liestal", get_temperature("GME00127786"))) %>%
  bind_rows(tibble(location = "kyoto", get_temperature("JA000047759"))) %>%
  bind_rows(tibble(location = "vancouver", get_temperature("CA001108395")))

```

```{r}
# DAILY TEMP
get_temp_daily <- function(station_id = station_id, date_range = c("1930-01-01", "2023-03-30")) {
  
  # Get daily maximum temperature data, minimum temperature and precipitation data
  tmax <- ghcnd_search(stationid = station_id, var = c("tmax"), 
                        date_min = date_range[1], date_max = date_range[2])[[1]]
  tmin <- ghcnd_search(stationid = station_id, var = c("tmin"), 
                        date_min = date_range[1], date_max = date_range[2])[[1]]
  prcp <- ghcnd_search(stationid = station_id, var = c("prcp"), 
                        date_min = date_range[1], date_max = date_range[2])[[1]]
  # Merge with tmax
  weather <- merge(tmax, tmin, by = "date") %>% merge(prcp, by = "date")
  
  # Select and transform relevant columns
  weather <- weather %>% 
    select(date, tmax, tmin, prcp) %>%
    mutate(tmax = tmax/10, tmin = tmin/10, # Convert temperature to tenth of the C
           year = as.integer(format(date, "%Y")),
           month = as.integer(strftime(date, '%m')) %% 12, # Make December "0"
           season = cut(month, breaks = c(0, 2, 5, 8, 11),
                        include.lowest = TRUE,
                        labels = c("Winter", "Spring", "Summer", "Fall")),
           year = if_else(month == 0, year + 1L, year)) %>%
    mutate(month_name = if_else(month == 0, 12, month)) %>%
    mutate(month_name = factor(month.name[as.integer(month_name)], levels = month.name)) 

  # Return transformed data
  return(weather)
}
```

```{r}
#Get data daily
daily_temperatures <-
  tibble(location = "washingtondc", get_temp_daily("USC00186350")) %>%
  bind_rows(tibble(location = "liestal", get_temp_daily("GME00127786"))) %>%
  bind_rows(tibble(location = "kyoto", get_temp_daily("JA000047759"))) %>%
  bind_rows(tibble(location = "vancouver", get_temp_daily("CA001108395")))


#Export file 
write.csv(daily_temperatures, "daily_temperatures_impute_missing.csv", row.names = FALSE)
```


## Python: Using KNN to impute misisng value
```{r}
#File can be found in cherry_imputing_missing.ipnyb
imputed_data <- read_csv("imputed_data.csv")

```

```{r}
# Accumulative growing days: GDD=(Tmax+Tmin)/2-Tbase
imputed_data <- imputed_data %>% 
  mutate(gdd= (tmax+tmin)/2 - 0)
View(imputed_data)

```

### Daily temperature would create too much noise for data -> using monthy temperature to model
```{r}
monthly_temp <- imputed_data %>%
  group_by(location, year, month, season) %>%
  summarize(tmax_avg = mean(tmax, na.rm = TRUE),
            tmax_max = max(tmax, na.rm = TRUE),
            tmin_avg = mean(tmin, na.rm = TRUE),
            tmin_min = min(tmin, na.rm = TRUE),
            gdd_avg = mean(gdd, na.rm = TRUE),
            prcp_avg = mean(prcp, na.rm = TRUE))


daily_temperatures%>% 
  group_by(location) %>% 
  slice_head(n = 5) 


```

```{r}
yearly_temp <- monthly_temp %>%
  filter(year >= 1960) %>%
  select(location, year, tmax_avg, tmin_avg, gdd_avg) %>%
  group_by(location, year) %>%
  summarise_each(funs(mean=mean(., na.rm=TRUE)))

# First, define colors.
BROWN <- "#AD8C97"
BROWN_DARKER <- "#7d3a46"
GREEN <- "#00FF00"
BLUE <- "#076FA1"
GREY <- "#C7C9CB"
GREY_DARKER <- "#5C5B5D"
RED <- "#E3120B"
  

plt1 <- yearly_temp %>%
  filter(year >= 1960) %>%
  ggplot(aes(year, gdd_avg_mean)) +
  geom_line(aes(color = location), size = 1, alpha=0.6) +
  geom_point(
    aes(fill = location), 
    pch = 21, # Type of point that allows us to have both color (border) and fill.
    color = "white", 
    stroke = 1 # The width of the border, i.e. stroke.
  ) +
  # Set values for the color and the fill
  scale_color_manual(values = c(BLUE, GREEN, BROWN, BROWN_DARKER)) +
  scale_fill_manual(values = c(BLUE, GREEN, BROWN, BROWN_DARKER)) +
  ggtitle("Yearly average temperature across 4 sites") +
  xlab("Year") + 
  ylab("Average Temperature")


plt1 <- plt1 + 
  scale_x_continuous(
    limits = c(1960, 2022)) +
  scale_y_continuous(
    limits = c(0, 32),
    breaks = seq(0, 30, by = 5), 
    expand = c(0, 0)
  ) + 
  theme(
    # Set background color to white
    panel.background = element_rect(fill = "white"),
    # Remove all grid lines
    panel.grid = element_line(colour = "grey50")
  )

plt1
```
_____________________________________________________________________
## Washington DC

```{r}
#import bloom date data
washington_bloom <-  cherry %>%
  filter(year >= 1960) %>%
  filter(location == "washingtondc")


#Filtering washington dc temperature
washington_temp <- monthly_temp %>% 
  filter(location == "washingtondc")


```

```{r}
# Get label for bloom date WASHINGTON - TESTING 
max(washington_bloom$bloom_doy) - min(washington_bloom$bloom_doy)

# Since prediction day can be +- 3 days, we create a category to observe how bloom date differs through years
washington_bloom$bloom_doy <- as.integer(washington_bloom$bloom_doy) 
wash_bloom <- washington_bloom %>% 
  select(lat, long, year, bloom_date, bloom_doy) %>%
  mutate(cat_bloom_doy = cut(bloom_doy, breaks = 8, labels = c(0,1,2,3,4,5,6,7))) 
head(wash_bloom)
```

```{r}
# join the datasets based on the "year" column
merged_washington <-left_join(wash_bloom, washington_temp, by='year')
# view the merged dataset
View(merged_washington)

sapply(merged_washington, function(x) sum(is.na(x)))

#Export file 
write.csv(merged_washington, "merged_washington.csv", row.names = FALSE)
```


____________________________________________________________________________________________________
## Kyoto
```{r}
#import bloom date data
kyoto_bloom <-  cherry %>%
  filter(year >= 1960) %>%
  filter(location == "kyoto")


#Filtering kyoto temperature
kyoto_temp <- monthly_temp %>% 
  filter(location == "kyoto")
```

```{r}
# Get label for bloom date KYOTO - TESTING 
max(kyoto_bloom$bloom_doy) - min(kyoto_bloom$bloom_doy)

# Since prediction day can be +- 4 days, we create a category to observe how bloom date differs through years
kyoto_bloom$bloom_doy <- as.integer(kyoto_bloom$bloom_doy) 
kyoto_bloom <- kyoto_bloom %>% 
  select(lat, long, year, bloom_date, bloom_doy) %>%
  mutate(cat_bloom_doy = cut(bloom_doy, breaks = 6, labels = c(0,1,2,3,4,5))) 
head(kyoto_bloom)
```

```{r}
# join the datasets based on the "year" column
merged_kyoto <-left_join(kyoto_bloom, kyoto_temp, by='year')
# view the merged dataset
View(merged_kyoto)

sapply(merged_kyoto, function(x) sum(is.na(x)))

#Export file 
write.csv(merged_kyoto, "merged_kyoto.csv", row.names = FALSE)
```

____________________________________________________________________________________________________
## Liestal
```{r}
#import bloom date data
liestal_bloom <-  cherry %>%
  filter(year >= 1960) %>%
  filter(location == "liestal")


#Filtering liestal temperature
liestal_temp <- monthly_temp %>% 
  filter(location == "liestal")
```

```{r}
# Get label for bloom date LIESTAL - TESTING 
max(liestal_bloom$bloom_doy) - min(liestal_bloom$bloom_doy)

# Since prediction day can be +- 4 days, we create a category to observe how bloom date differs through years
liestal_bloom$bloom_doy <- as.integer(liestal_bloom$bloom_doy) 
liestal_bloom <- liestal_bloom %>% 
  select(lat, long, year, bloom_date, bloom_doy) %>%
  mutate(cat_bloom_doy = cut(bloom_doy, breaks = 11, labels = c(0,1,2,3,4,5,6,7,8,9,10))) 
head(liestal_bloom)
```

```{r}
# join the datasets based on the "year" column
merged_liestal <-left_join(liestal_bloom, liestal_temp, by='year')
# view the merged dataset
View(merged_liestal)

sapply(merged_liestal, function(x) sum(is.na(x)))

#Export file 
write.csv(merged_liestal, "merged_liestal.csv", row.names = FALSE)
```


____________________________________________________________________________________________________
## Vancouver
```{r}
#import bloom date data
vancouver_bloom <-  cherry %>%
  filter(year >= 1960) %>%
  filter(location == "vancouver")


#Filtering vancouver temperature
vancouver_temp <- monthly_temp %>%
  filter(location == "vancouver") %>%
  filter(year >= 1960) 
```

```{r}
# join the datasets based on the "year" column
merged_vancouver <-left_join(vancouver_temp,vancouver_bloom, by='location')
# view the merged dataset
View(merged_vancouver)

sapply(merged_vancouver, function(x) sum(is.na(x)))

#Export file
write.csv(merged_vancouver, "merged_vancouver.csv", row.names = FALSE)
```

### Adding 2 features to the dataset for modelling: 

